---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(patchwork)
source("scripts/theme_black.R")
```

All Uniprot AMPs were downloaded from UniProt on 07 July 2021 
```{r}
uniprot_amps <- read_tsv("data/uniprot_amps_07Jul21_3371rev42126unrev.tab.gz") %>%
  rename(Entry_name = `Entry name`) %>% mutate(Organism = str_remove(Organism, " \\(.*")) %>% 
  rename(Taxonomic_lineage = `Taxonomic lineage (ALL)`) %>% 
  rename(Order = `Taxonomic lineage (ORDER)`) %>% 
  rename(Phylum = `Taxonomic lineage (PHYLUM)`) %>%
  rename(Class = `Taxonomic lineage (CLASS)`) %>%
  mutate(Order = str_remove(Order, " \\(.*")) %>%
  mutate(Organism = str_replace_all(Organism, " ", "_")) %>% 
  filter(!grepl("Viruses", Taxonomic_lineage)) %>% 
  filter(!grepl("unclassified", Taxonomic_lineage)) %>%
  mutate(label = case_when(str_detect(Taxonomic_lineage, "Bacteria") ~ "Bacteria",
                           str_detect(Taxonomic_lineage, "Archaea") ~ "Archaea",
                               str_detect(Phylum, "Nematoda*") ~ "Nematoda",
                               str_detect(Phylum, "Foraminifera*") ~ "Foraminifera",
                               str_detect(Phylum, "Bryozoa*") ~ "Bryozoa",
                               str_detect(Phylum, "Brachiopoda*") ~ "Brachiopoda",
                               str_detect(Phylum, "Rotifera*") ~ "Rotifera",
                               str_detect(Phylum, "Tardigrada*") ~ "Tardigrada",
                               str_detect(Phylum, "Porifera*") ~ "Porifera",
                                               TRUE ~ Phylum)) %>%
  mutate(Order = case_when(
    str_detect(Organism, "Lates_calcarifer") ~ "Perciformes",
    str_detect(Organism, "Parambassis_ranga") ~ "Perciformes",
    str_detect(Organism, "Larimichthys_crocea") ~ "Acanthuriformes",
    str_detect(Organism, "Amphiprion") ~ "Perciformes",
    str_detect(Organism, "Collichthys_lucidus") ~ "Perciformes",
    str_detect(Organism, "Siganus_canaliculatus") ~ "Perciformes",
    str_detect(Organism, "Parambassis_ranga") ~ "Perciformes",
    str_detect(Organism, "Miichthys_miiuy") ~ "Perciformes",
    str_detect(Organism, "Dicentrarchus_labrax") ~ "Perciformes",
    str_detect(Organism, "Scatophagus_argus") ~ "Perciformes",
    str_detect(Organism, "Biomphalaria_glabrata") ~ "Hygrophila",
    str_detect(Organism, "Stegastes") ~ "Perciformes",
    str_detect(Organism, "Chrysochloris_asiatica") ~ "Afrosoricida",
    str_detect(Organism, "Totoaba_macdonaldi") ~ "Acanthuriformes",
    str_detect(Organism, "Acanthochromis_polyacanthus") ~ "Perciformes",
    str_detect(Organism, "Argyrosomus_regius") ~ "Perciformes",
    str_detect(Organism, "Collichthys_lucidus") ~ "Perciformes",
    str_detect(Organism, "Morone") ~ "Perciformes",
    str_detect(Organism, "Capitella_teleta") ~ "Capitellida",
    str_detect(Organism, "Arenicola_marina") ~ "Capitellida",
    str_detect(Organism, "Naegleria_fowleri") ~ "Schizopyrenida",
    str_detect(Organism, "Dimorphilus_gyrociliatus") ~ "Eunicida",
    str_detect(Organism, "Reticulomyxa_filosa") ~ "Athalamida",
    TRUE ~ Order)) %>%
     mutate(Class = case_when(str_detect(Class, "Lepidosauria*") ~ "Reptilia",
                           str_detect(Order, "Crocodylia") ~ "Reptilia",
                           str_detect(Order, "Testudines") ~ "Reptilia",
                                               TRUE ~ Class))
```

## Taxonomic level: Phylum


```{r, fig.width=10, fig.height=6, echo = FALSE}
uniprot_amps %>%
  filter(!grepl("Bacteria", label)) %>% 
  filter(!grepl("Archaea", label)) %>%
  group_by(label) %>% 
  summarise(AMP_count = n()) %>%
  mutate(percentage = round(AMP_count / sum(AMP_count) * 100, digits = 2)) %>%
  arrange(.by_group = TRUE, desc(AMP_count)) %>%
  ggplot(aes(x = reorder(label, AMP_count), y = AMP_count)) +
  geom_col(width = 0.7, fill = "cyan") +
  coord_flip() +
  geom_text(aes(label=paste0(percentage, "%")), size = 2.4, hjust = -0.1, colour = "white") +
  labs(x = "Taxonomic Phyla", y = "Number of antimicrobial peptides (AMPs) in UniProt ", fill = "") +
  theme_black() +
  theme(legend.position = "bottom")

ggsave("figures/phyla_amps.png", width = 10, height = 4, dpi = 300)
```

**Figure 1.1:** The number of antimicrobial peptides in phyla across the UniProt database (excluding Bacteria, Viruses and Archaea)

## Taxonomic level: Class

```{r, fig.width=10, fig.height=6, echo = FALSE}
chordata_amps_plot <- uniprot_amps %>%
  filter(Phylum == "Chordata") %>%
  group_by(Class) %>% 
  summarise(AMP_count = n()) %>%
  mutate(percentage = round(AMP_count / sum(AMP_count) * 100, digits = 2)) %>%
  arrange(.by_group = TRUE, desc(AMP_count)) %>%
  ggplot(aes(x = reorder(Class, AMP_count), y = AMP_count)) +
  geom_col(width = 0.7, fill = "cyan") +
  coord_flip() +
  #geom_text(aes(label=paste0(percentage, "%")), size = 2.4, hjust = -0.1, colour = "white") +
  labs(x = "Taxonomic Class", y = "AMP count in Chordata", fill = "") +
  theme_black() 

arth_amps <- uniprot_amps %>% filter(Phylum == "Arthropoda") %>%
       mutate(Class = case_when(str_detect(Class, "Collembola*") ~ "Collembola",
                           str_detect(Class, "Merostomata*") ~ "Merostomata",
                           str_detect(Class, "Chilopoda") ~ "Chilopoda",
                                               TRUE ~ Class))

arthropoda_amps_plot <- arth_amps %>%
  group_by(Class) %>% 
  summarise(AMP_count = n()) %>%
  mutate(percentage = round(AMP_count / sum(AMP_count) * 100, digits = 2)) %>%
  arrange(.by_group = TRUE, desc(AMP_count)) %>%
  ggplot(aes(x = reorder(Class, AMP_count), y = AMP_count)) +
  geom_col(width = 0.7, fill = "cyan") +
  coord_flip() +
  #geom_text(aes(label=paste0(percentage, "%")), size = 2.4, hjust = -0.1, colour = "white") +
  labs(x = "Taxonomic Class", y = "AMP count in Arthropoda", fill = "") +
  theme_black() 


chordata_amps_plot | arthropoda_amps_plot

ggsave(plot = chordata_amps_plot, "figures/chor_class_amps.png", width = 5, height = 3)
ggsave(plot = arthropoda_amps_plot, "figures/arth_class_amps.png", width = 5, height = 3)
```

**Figure 1.2:** The number of antimicrobial peptides in classes in chordates (left) and Arthropods (right) 


## Taxonomic level: Order

```{r, fig.width=10, fig.height=6, echo = FALSE}

mammal_amps_plot <- uniprot_amps %>% filter(grepl("Mammalia", Taxonomic_lineage)) %>% 
  group_by(Order) %>% 
  summarise(AMP_count = n()) %>%
  arrange(.by_group = TRUE, desc(AMP_count)) %>%
  pivot_longer(cols = starts_with("AMP")) %>%
  ggplot(aes(x = reorder(Order, value), y = value)) +
  geom_col(width = 0.7, fill = "cyan") +
  coord_flip() +
  labs(x = "Order", y = "Number of AMPs described in mammals", fill = "") +
  theme_classic() +
  theme(legend.position = "bottom") +
  theme_black() 

insect_amps_plot <- uniprot_amps %>% filter(grepl("Insecta", Taxonomic_lineage)) %>% 
  group_by(Order) %>% 
  summarise(AMP_count = n()) %>%
  arrange(.by_group = TRUE, desc(AMP_count)) %>%
  pivot_longer(cols = starts_with("AMP")) %>%
  ggplot(aes(x = reorder(Order, value), y = value)) +
  geom_col(width = 0.7, fill = "cyan") +
  coord_flip() +
  labs(x = "Order", y = "Number of AMPs described in insects", fill = "") +
  theme_classic() +
  theme(legend.position = "bottom") +
  theme_black() 

mammal_amps_plot | insect_amps_plot

ggsave(plot = mammal_amps_plot, "figures/mammal_order_amps.png", width = 5, height = 3)
ggsave(plot = insect_amps_plot, "figures/insect_order_amps.png", width = 5, height = 3)

```

## Taxonomic level: Species

```{r, fig.width=12, fig.height=6, echo = FALSE}
mammal_orders <- c("Primates", "Artiodactyla", "Carnivora", "Rodentia", "Chiroptera")

mammal_selection <- uniprot_amps %>% filter(grepl("Mammalia", Taxonomic_lineage)) %>% filter(Order %in% mammal_orders)

mammal_species_amps_plot <- mammal_selection %>%
  group_by(Organism, Order) %>% 
  summarise(AMP_count = n(), .groups = "drop" ) %>%
  group_by(Order) %>%
  slice_max(AMP_count, n = 3) %>%
  ggplot(aes(x = reorder(Organism, AMP_count), y = AMP_count)) +
  geom_col(width = 0.7, fill = "cyan") +
  coord_flip() +
  facet_wrap(~Order, ncol = 1, scales = "free_y") +
  labs(x = "Mammal species (Top 3)", y = "Number of antimicrobial peptides (AMPs)", fill = "") +
  theme_black() +
  theme(axis.text.y = element_text(face = "italic"))


ggsave(plot = mammal_species_amps_plot, "figures/mammal_species_amps.png", width = 7, height = 6)


insect_orders <- c("Diptera", "Lepidoptera", "Hymenoptera", "Coleoptera", "Hemiptera")
insect_selection <- uniprot_amps %>% filter(grepl("Insecta", Taxonomic_lineage)) %>% filter(Order %in% insect_orders)

insect_species_amps_plot <- insect_selection %>%
  group_by(Organism, Order) %>% 
  summarise(AMP_count = n(), .groups = "drop" ) %>%
  group_by(Order) %>%
  slice_max(AMP_count, n = 3) %>%
  ggplot(aes(x = reorder(Organism, AMP_count), y = AMP_count)) +
  geom_col(width = 0.7, fill = "cyan") +
  coord_flip() +
  facet_wrap(~Order, ncol = 1, scales = "free_y") +
  labs(x = "Insect species (Top 3)", y = "Number of antimicrobial peptides (AMPs)", fill = "") +
  theme_black() +
  theme(axis.text.y = element_text(face = "italic"))

ggsave(plot = insect_species_amps_plot, "figures/insect_species_amps.png", width = 7, height = 6)

mammal_species_amps_plot | insect_species_amps_plot
```

**Figure 1.3:** The number of antimicrobial peptides in mammal species (left) and insect species (right) 

```{r, echo = FALSE, eval = FALSE}
m1 <- uniprot_amps %>% filter(grepl("Mammalia", Taxonomic_lineage)) %>% 
  group_by(Order) %>% 
  summarise(AMP_count = n()) %>%
  arrange(.by_group = TRUE, desc(AMP_count)) %>%
  pivot_longer(cols = starts_with("AMP")) %>%
  ggplot(aes(x = reorder(Order, value), y = value)) +
  geom_col(width = 0.7, fill = "forestgreen") +
  coord_flip() +
  labs(x = "Mammalian Order", y = "Number of described AMPs", fill = "") +
  theme_classic() +
  theme(legend.position = "bottom")

m2 <- mammal_selection %>%
  group_by(Organism, Order) %>% 
  summarise(AMP_count = n(), .groups = "drop" ) %>%
  group_by(Order) %>%
  slice_max(AMP_count, n = 3) %>%
  ggplot(aes(x = reorder(Organism, AMP_count), y = AMP_count)) +
  geom_col(width = 0.7, fill = "forestgreen") +
  coord_flip() +
  facet_wrap(~factor(Order, levels = c("Primates", "Artiodactyla", "Carnivora", "Rodentia", "Chiroptera")), ncol = 1, scales = "free_y") +
  labs(x = "Mammal species (Top three)", y = "Number of described AMPs", fill = "") +
  theme_classic() +
  theme(axis.text.y = element_text(face = "italic"),
        strip.background = element_rect(fill = NA, color = NA),
        strip.text.x = element_text(size = 9, face = "bold"))

m1 | m2 + plot_annotation(tag_levels = "A") 

ggsave("figures/mammals.png", width = 10, height = 6)
```


```{r, echo = FALSE}
uniprot_amps %>% 
  filter(grepl("Amphibia", Taxonomic_lineage)) %>% 
  group_by(Organism, Order) %>% 
  summarise(AMP_count = n(), .groups = "drop" ) %>%
  group_by(Order) %>%
  slice_max(AMP_count, n = 20) %>%
  filter(Order == "Anura") %>%
  ggplot(aes(x = reorder(Organism, AMP_count), y = AMP_count)) +
  geom_col(width = 0.7, fill = "cyan") +
  coord_flip() +
  labs(x = "Anuran species (Top 20)", y = "Number of antimicrobial peptides (AMPs)", fill = "") +
  theme_black() +
  theme(axis.text.y = element_text(face = "italic"))

ggsave("figures/anuran_amps.png", width = 7, height = 3)

```

**Figure 1.4:** The number of antimicrobial peptides in frogs and toads (top 20)


## Fin


```{r}
uniprot_amps %>% filter(grepl("Mammalia", Taxonomic_lineage)) %>% 
  group_by(Organism) %>% 
  summarise(AMP_count = n()) %>%
  arrange(.by_group = TRUE, desc(AMP_count)) %>%
  slice_max(order_by = AMP_count, n = 30) %>%
  ggplot(aes(x = reorder(Organism, AMP_count), y = AMP_count)) +
  geom_bar(stat ="identity", position = "dodge") +
  coord_flip() +
  labs(x = "Mammal species (Top 100)", y = "Number of antimicrobial peptides (AMP)", fill = "") +
  theme_classic() +
  theme(axis.text.y = element_text(face = "italic"))
```






Similar to the reviewed AMPs, the orders Primates, Artiodactyla dominate the top three orders that have the highest AMP count. However, in the unreviewed AMPs, Carnivora has topped Rodentia in number of AMPs. In addition, Chiroptera also appears in the top 5 orders that have the most number of AMPs in TrEMBL. 

Within Carnivora, the domestic dog, [*Canis lupus familiaris*](https://www.uniprot.org/proteomes/UP000002254) contains the most AMPs (72). In Chiroptera, the greater horseshoe bat, [*Rhinolophus ferrumequinum*](https://www.uniprot.org/proteomes/UP000472240) contains the most AMPs (54). As these organisms contain the most AMPs for their respective orders, and both contain a reference proteome, these two species could be good candidates to include in the analysis.


*NOTE: problem with this though as these are in the unreviewed section, and not well represented in the reviewed AMPs (used for the training and query data) at all. How to get around this? use unreviewed AMPs instead for training / query data? or just do normal approach and then these 2 organisms can be used as a decent test for taxonomic bias in training/query data  ... (and still have decent number of AMPs to test effectiveness of the model/BLAST methods )*

```{r}
uniprot_amps %>% filter(grepl("Carnivora", Taxonomic_lineage))  %>% count(Organism, sort = TRUE) %>% slice_head(n=6)

uniprot_amps %>% filter(grepl("Chiroptera", Taxonomic_lineage)) %>% count(Organism, sort = TRUE) %>% slice_head(n=6)
```

Marsupials and Monotremes

```{r}
uniprot_amps %>% filter(grepl("Metatheria", Taxonomic_lineage)) %>% count(Organism, sort = TRUE)

uniprot_amps %>% filter(grepl("Monotremata", Taxonomic_lineage)) %>% count(Organism, sort = TRUE)
```
