---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(ape)
library(treeio)
library(patchwork)
```

## Prepare AMP and tree datasets


AMPs (3,371 sequences) were obtained from SwissProt on 01 July 2021 [UniProtKB 2021_03 results](https://www.uniprot.org/uniprot/?query=keyword%3A%22Antimicrobial%20%5BKW-0929%5D%22%20AND%20reviewed%3Ayes&columns=id%2Centry%20name%2Creviewed%2Cprotein%20names%2Cgenes%2Corganism%2Clength%2Ckeyword-id%2Ckeywords%2Cproteome%2Corganism-id%2Clineage(ORDER)%2Csequence%2Cexistence%2Clineage(ALL)&sort=sequence-modified). 


Read in the same SwissProt AMPs used for creating the classification models and BLAST datasets. 
The organism names were tidied up and made more compatible to [TimeTree](http://timetree.org/) names where species only have two names.
Viruses were removed. This resulted in 802 unique organisms. Species not fully defined (e.g. "Lactococcus sp) were additionally removed. This resulted in 769 species.

```{r}
swissprot_amps <- read_tsv("data/uniprot-keywordAntimicrobial+[KW-0929]-filtered-reviewedyes24May21.tab") %>% 
  rename(Entry_name = `Entry name`) %>% 
  rename(Taxonomic_lineage = `Taxonomic lineage (ALL)`) %>% 
  filter(!grepl("Viruses", Taxonomic_lineage)) %>%
  filter(!grepl("\\.", Organism)) %>%
  mutate(Organism = word(Organism, 1, 2)) %>% 
  mutate(Organism = case_when(
    str_detect(Organism, "Phlyctimantis maculatus") ~ "Hylambates maculatus",
    str_detect(Organism, "Nyctimystes infrafrenatus") ~ "Litoria infrafrenata",
    str_detect(Organism, "Lithobates catesbeianus") ~ "Rana catesbeiana",
    str_detect(Organism, "Boana punctata") ~ "Hypsiboas punctatus",
                          TRUE ~ Organism)) %>%
  mutate(Organism = str_replace(Organism, "Ranoidea", "Litoria")) %>%
  mutate(Organism = str_replace_all(Organism, " ", "_")) 

write_lines(unique(swissprot_amps$Organism), "cache/organism_list.txt")
```

Add in the unreviewed horseshoe bat AMPs (used in previous analysis)
```{r}
horseshoe_bat_amps <- read_tsv("data/uniprot_amps_07Jul21_3371rev42126unrev.tab.gz") %>%
  rename(Entry_name = `Entry name`) %>% 
  rename(Taxonomic_lineage = `Taxonomic lineage (ALL)`) %>%
  mutate(Organism = str_remove(Organism, " \\(.*")) %>% 
  mutate(Organism = str_replace_all(Organism, " ", "_")) %>%
  filter(Organism == "Rhinolophus_ferrumequinum") %>%
  select(!c(`Taxonomic lineage (PHYLUM)`, `Taxonomic lineage (CLASS)`, `Organism ID`))

swissprot_amps <- rbind(swissprot_amps, horseshoe_bat_amps)
```


Read in the tree from [TimeTree](http://timetree.org/)

Normalise names:
211 organisms had unresolved names of which 76 (mostly bacteria) were replaced with different names and the remaining 135 organisms potentially were not in the TimeTree database at the time. 


```{r}
timetree <- read.tree("data/organism_list.nwk")
timetree <- read.tree("data/organism_list_w_bat.nwk")

timetree_tibble <- as_tibble(timetree) %>%
   mutate(label = case_when(
    str_detect(label, "Halobacterium_salinarum") ~ "Haloarchaeon_S8a",
    str_detect(label, "_aerogenes") ~ "Klebsiella_pneumoniae",
    str_detect(label, "Achromobacter_denitrificans") ~ "Achromobacter_lyticus",
    str_detect(label, "Lactobacillus_casei") ~ "Lactobacillus_paracasei",
    str_detect(label, "Actinoplanes_missouriensis") ~ "Actinoplanes_garbadinensis",
    str_detect(label, "Agrocybe_praecox") ~ "Agrocybe_cylindracea",
    str_detect(label, "Coprinopsis_lagopus") ~ "Coprinopsis_cinerea",
    str_detect(label, "Ganoderma_sinense") ~ "Pleurotus_sajor-caju",
    str_detect(label, "Pseudoplectania_nigrella") ~ "Peziza_vesiculosa",
    str_detect(label, "Pseudallescheria_apiosperma") ~ "Microascus_cirrosus",
    str_detect(label, "Colletotrichum_acutatum") ~ "Colletotrichum_dematium",
    str_detect(label, "Penicillium_javanicum") ~ "Penicillium_rubens",
    str_detect(label, "Mucor_racemosus") ~ "Rhizomucor_pusillus",
    str_detect(label, "Schistosoma_japonicum") ~ "Schistosoma_mansoni",
    str_detect(label, "Molgula_tectiformis") ~ "Halocynthia_aurantium",
    str_detect(label, "Paralichthys_dentatus") ~ "Paralichthys_olivaceus",
    str_detect(label, "Siganus_vulpinus") ~ "Siganus_canaliculatus",
    str_detect(label, "Opsanus_tau") ~ "Thalassophryne_nattereri",
    str_detect(label, "Paracentrotus_lividus") ~ "Echinus_esculentus",
    str_detect(label, "Leptasterias_hexactis") ~ "Asterias_rubens",
    str_detect(label, "Scolopendra_cingulata") ~ "Scolopendra_subspinipes",
    str_detect(label, "Panulirus_interruptus") ~ "Panulirus_argus",
    str_detect(label, "Chionoecetes_opilio") ~ "Hyas_araneus",
    str_detect(label, "Anax_junius") ~ "Aeshna_cyanea",
    str_detect(label, "Anoplius_aethiops") ~ "Anoplius_samariensis",
    str_detect(label, "Osmia_lignaria") ~ "Osmia_rufa",
    str_detect(label, "Lasioglossum_calceatum") ~ "Lasioglossum_laticeps",
    str_detect(label, "Macropis_europaea") ~ "Macropis_fulvipes",
    str_detect(label, "Tetramorium_caespitum") ~ "Tetramorium_bicarinatum",
    str_detect(label, "Formica_moki") ~ "Formica_aquilonia",
    str_detect(label, "Odontomachus_clarus") ~ "Odontomachus_monticola",
    str_detect(label, "Polistes_tenebricosus") ~ "Polistes_hebraeus",
    str_detect(label, "Stizoides_foxi") ~ "Sphecius_speciosus ",
    str_detect(label, "Aleiodessp.BOLD") ~ "Pimpla_disparis",
    str_detect(label, "Osmoderma_eremita") ~ "Trypoxylus_dichotomus",
    str_detect(label, "Tetraopes_tetrophthalmus") ~ "Acalolepta_luxuriosa",
    str_detect(label, "Chrysolina_hyperici") ~ "Gastrophysa_atrocyanea",
    str_detect(label, "Phlebotomus_papatasi") ~ "Phlebotomus_duboscqi",
    str_detect(label, "Sarcophaga_bullata") ~ "Sarcophaga_peregrina",
    str_detect(label, "Tinea_columbariella") ~ "Oiketicus_kirbyi",
    str_detect(label, "Manga_basilinea") ~ "Mamestra_brassicae",
    str_detect(label, "Heliothis_terracottoides") ~ "Heliothis_virescens",
    str_detect(label, "Saturnia_mendocino") ~ "Antheraea_mylitta",
    str_detect(label, "Accinctapubes_albifasciata") ~ "Galleria_mellonella",
    str_detect(label, "Aeschyntelus_notatus") ~ "Riptortus_clavatus",
    str_detect(label, "Brochymena_sp._WCW-2003") ~ "Podisus_maculiventris ",
    str_detect(label, "Platypleura_capensis") ~ "Cicada_flammata",
    str_detect(label, "Cryptotympana_atrata") ~ "Cryptotympana_dubia",
    str_detect(label, "Acyrthosiphon_pisum") ~ "Megoura_viciae",
    str_detect(label, "Ixodes_hexagonus") ~ "Ixodes_sinensis",
    str_detect(label, "Rhipicephalus_sanguineus") ~ "Rhipicephalus_microplus",
    str_detect(label, "Ornithodoros_moubata") ~ "Argas_monolakensis",
    str_detect(label, "Loxosceles_laeta") ~ "Loxosceles_gaucho",
    str_detect(label, "Lycosa_tarantula") ~ "Oxyopes_takobius",
    str_detect(label, "Vaejovis_bandido") ~ "Mesomexovis_subcristatus",
    str_detect(label, "Metaphire_sieboldi") ~ "Metaphire_guillelmi",
    str_detect(label, "Poecilobdella_manillensis") ~ "Hirudo_medicinalis",
    str_detect(label, "Nereis_denhamensis") ~ "Perinereis_aibuhitensis",
    str_detect(label, "Capitella_teleta") ~ "Arenicola_marina",
    str_detect(label, "Conus_purpurascens") ~ "Conus_mustelinus",
    str_detect(label, "Globba_radicalis") ~ "Curcuma_longa",
    str_detect(label, "Xanthosoma_helleborifolium") ~ "Xanthosoma_sagittifolium",
    str_detect(label, "Macadamia_tetraphylla") ~ "Macadamia_integrifolia",
    str_detect(label, "Brassica_rapa") ~ "Brassica_campestris",
    str_detect(label, "Cucurbita_pepo") ~ "Cucurbita_maxima",
    str_detect(label, "Inga_cf._edulis_Klitgaard_677") ~ "Inga_vera",
    str_detect(label, "Parkia_timoriana") ~ "Parkia_pendula",
    str_detect(label, "Cullen_australasicum") ~ "Cullen_corylifolium",
    str_detect(label, "Arachis_major") ~ "Arachis_hypogaea",
    str_detect(label, "Heuchera_merriamii") ~ "Heuchera_sanguinea",
    str_detect(label, "Portulaca_grandiflora") ~ "Basella_alba",
    str_detect(label, "Amaranthus_palmeri") ~ "Amaranthus_caudatus ",
    str_detect(label, "Bidens_cernua") ~ "Dahlia_merckii",
    str_detect(label, "Geophila_obvallata") ~ "Chassalia_chartacea ",
    str_detect(label, "Psychotria_kirkii") ~ "Psychotria_longipes",
    str_detect(label, "Lippia_javanica") ~ "Lippia_sidoides",
    str_detect(label, "Diospyros_cavalcantei") ~ "Diospyros_texana",
    str_detect(label, "_gnavus") ~ "Ruminococcus_gnavus",
    str_detect(label, "Mycobacterium_phlei") ~ "Mycolicibacterium_phlei",
    str_detect(label, "Kassina_maculata") ~ "Hylambates_maculatus",
    str_detect(label, "Rana_septentrionalis") ~ "Lithobates_septentrionalis",
    str_detect(label, "Rana_sevosa") ~ "Lithobates_sevosus",
    str_detect(label, "Rana_clamitans") ~ "Lithobates_clamitans",
    str_detect(label, "Rana_berlandieri") ~ "Lithobates_berlandieri",
    str_detect(label, "Rana_sphenocephala") ~ "Lithobates_sphenocephalus",
    str_detect(label, "Rana_palustris") ~ "Lithobates_palustris",
    str_detect(label, "Rana_pipiens") ~ "Lithobates_pipiens",
    str_detect(label, "Rana_sylvatica") ~ "Lithobates_sylvaticus",
    str_detect(label, "Rugosa_rugosa") ~ "Nidirana_pleuraden",
    str_detect(label, "Babina_okinavana") ~ "Glandirana_rugosa",
    str_detect(label, "Babina_pleuraden") ~ "Nidirana_pleuraden",
    str_detect(label, "Eupemphix_nattereri") ~ "Physalaemus_nattereri",
    str_detect(label, "Hypsiboas_albopunctatus") ~ "Boana_albopunctata",
    str_detect(label, "Phyllomedusa_sauvagii") ~ "Phyllomedusa_sauvagei",
    str_detect(label, "Phyllomedusa_hypochondrialis") ~ "Pithecopus_hypochondrialis",
    str_detect(label, "Phyllomedusa_oreades") ~ "Pithecopus_oreades",
    str_detect(label, "Pachymedusa_dacnicolor") ~ "Agalychnis_dacnicolor",
    str_detect(label, "Hylomantis_lemur") ~ "Agalychnis_lemur",
    str_detect(label, "Cercopithecus_preussi") ~ "Allochrocebus_preussi",
    str_detect(label, "Fenneropenaeus_merguiensis") ~ "Penaeus_merguiensis",
    str_detect(label, "Litopenaeus_vannamei") ~ "Penaeus_vannamei",
    str_detect(label, "Litopenaeus_setiferus") ~ "Penaeus_setiferus",
    str_detect(label, "Neobellieria_bullata") ~ "Sarcophaga_peregrina",
                          TRUE ~ label)) %>%
  mutate(label = str_trim(label, side = "right")) # remove trailing whitespace
```


```{r}
treelabel_notinAMPs <- timetree_tibble[!timetree_tibble$label %in% swissprot_amps$Organism,]

AMPs_notintreelabel<- swissprot_amps[!swissprot_amps$Organism %in% timetree_tibble$label,]
```

After renaming the majority of species in the tree to match to the species names used in SwissProt, 456 AMP entries remained in SwissProt where the species could not be found in the TimeTree database. These mostly includes arachnids (scorpions and spiders), hymenopterans (wasps, bees, ants and sawflies) and anurans (frogs).

```{r}
AMPs_notintreelabel %>% count(`Taxonomic lineage (ORDER)`, sort = TRUE)
```

Convert the tree tibble back to a tree
```{r}
timetree_reworded <- as.treedata(timetree_tibble)

timetree_reworded_phylo <- timetree_reworded@phylo
```

## Pairwise distances between organisms

Calculate pairwise distances between pairs of tips from the tree using the branch lengths with [`cophenetic.phylo`](https://rdrr.io/cran/ape/man/cophenetic.phylo.html)


```{r}
timetree_distance_matrix <- cophenetic.phylo(timetree_reworded_phylo)
```

Convert matrix to dataframe and left join to SwissProt AMPs dataset (by organism). Then calculate the inverse pairwise distance (1/distance) for each organism
```{r}
timetree_dm_df <- as.data.frame(timetree_distance_matrix) %>% rownames_to_column("Organism")

amps_w_distance <- swissprot_amps %>%
  left_join(timetree_dm_df, by = "Organism") %>% 
  mutate(across(.cols = Haloarchaeon_S8a:Cycas_revoluta, function(x) 1/x, .names = "{.col}_inverse" ))
```


```{r, echo = FALSE, fig.height=10}
distance_plot <- amps_w_distance %>% 
  select(Homo_sapiens, Pan_troglodytes, Mus_musculus , Rattus_norvegicus , Bos_taurus, Sus_scrofa, Canis_lupus, Rhinolophus_ferrumequinum) %>%
  pivot_longer(names_to = "Organism", values_to = "Distance", cols = everything()) %>%
  ggplot(aes(x = Distance)) +
  geom_histogram(bins = 40) +
  facet_wrap(~Organism, ncol = 2, scales = "free") +
  theme_classic() +
  theme(strip.text.x = element_text(face = "italic"),
        strip.background = element_rect(fill = "white", color = "white"))

inverse_distance_plot <- amps_w_distance %>% 
  select(Homo_sapiens_inverse, Pan_troglodytes_inverse, Mus_musculus_inverse , Rattus_norvegicus_inverse , Bos_taurus_inverse, Sus_scrofa_inverse, Canis_lupus_inverse, Rhinolophus_ferrumequinum_inverse) %>%
  pivot_longer(names_to = "Organism", values_to = "Distance_inverse", cols = everything()) %>%
  ggplot(aes(x = Distance_inverse)) +
  geom_histogram(bins = 40) +
  facet_wrap(~Organism, ncol = 2, scales = "free") +
  theme_classic() +
  theme(strip.text.x = element_text(face = "italic"),
        strip.background = element_rect(fill = "white", color = "white"))

distance_plot / inverse_distance_plot + plot_annotation(tag_levels = "A")

ggsave("figures/distancevsinversedistance.png", width = 8, height = 9)
```

**Figure 4.1:** **A** Histogram of pairwise distance **B** Histogram of the inverse pairwise distance between each faceted organism and 623 out of the 756 other organisms present in the AMP dataset

## Summed inverse pairwise distance vs. AUPRC 

Replace `inf` values to `NA` and sum inverse distance values. Then rename columns to original organism names and transform to longer format so it can be more easily combined to the AUPRC results dataframe  
```{r}

summed_inverse_distance <- amps_w_distance %>%
  map_df(function(x) replace(x, is.infinite(x), NA)) %>%
  summarise(across(.cols = c(Homo_sapiens_inverse, Pan_troglodytes_inverse, Mus_musculus_inverse , Rattus_norvegicus_inverse , Bos_taurus_inverse, Sus_scrofa_inverse, Canis_lupus_inverse, Rhinolophus_ferrumequinum_inverse), sum, na.rm = TRUE, .names = "{.col}_sum"))

summed_inverse_distance <- amps_w_distance %>%
  map_df(function(x) replace(x, is.infinite(x), NA)) %>%
  summarise(across(.cols = c(Homo_sapiens_inverse, Pan_troglodytes_inverse, Mus_musculus_inverse , Rattus_norvegicus_inverse , Bos_taurus_inverse, Sus_scrofa_inverse, Canis_lupus_inverse, Rhinolophus_ferrumequinum_inverse), sum, na.rm = TRUE, .names = "{.col}_sum")) %>% rename("Homo_sapiens" = Homo_sapiens_inverse_sum, "Pan_troglodytes" = Pan_troglodytes_inverse_sum, "Mus_musculus" = Mus_musculus_inverse_sum, "Rattus_norvegicus" = Rattus_norvegicus_inverse_sum, "Bos_taurus" = Bos_taurus_inverse_sum, "Sus_scrofa" = Sus_scrofa_inverse_sum, "Canis_lupus_familiaris" = Canis_lupus_inverse_sum,"Rhinolophus_ferrumequinum" = Rhinolophus_ferrumequinum_inverse_sum) %>% pivot_longer(cols = everything(), names_to = "Organism", values_to = "Inverse_distance_sum")

```

Read in previously calculated AUPRC values (see 03_blast_and_prediction.Rmd). 
Widen data so its easier to add the summed_inverse_distance dataframe to it

```{r}
methods_auprc <- readRDS("cache/methods_auprc.rds")

methods_auprc_wide <- pivot_wider(methods_auprc, names_from = Method, values_from = AUPRC)
```

Join the AUPRC values and distance metric for each organism
```{r}
auprc_and_distance_metric <- left_join(methods_auprc_wide, summed_inverse_distance, by = "Organism")
```

Change back to long format for plotting 
```{r}
auprc_and_distance_metric_long <- auprc_and_distance_metric %>% pivot_longer(cols = c(Classification, BLAST1, BLAST2), names_to = "Method", values_to = "AUPRC")
```

```{r, echo=FALSE}
ggplot(auprc_and_distance_metric_long, aes(x = Inverse_distance_sum, y = AUPRC, colour = Method)) +
  geom_line() +
  labs(x = "The sum of the inverse pairwise distance", colour = "") +
  theme_classic() +
  theme(legend.position = "bottom")

ggsave("figures/auprc_vs_distance.png", width = 7, height = 5)
```

**Figure 4.2:** Line plot of the summed inverse pairwise distance and the AUPRC for each AMP finding method 


*inverse values have a fair `inf` values*
```{r}
length(which(is.infinite(amps_w_distance$Homo_sapiens_inverse))) #99

length(which(is.infinite(amps_w_distance$Mus_musculus_inverse))) #100

length(which(is.infinite(amps_w_distance$Pan_troglodytes_inverse))) #38

length(which(is.infinite(amps_w_distance$Bos_taurus_inverse))) #55

length(which(is.infinite(amps_w_distance$Canis_lupus_inverse))) #6

length(which(is.infinite(amps_w_distance$Rhinolophus_ferrumequinum_inverse))) #54

length(which(is.infinite(amps_w_distance$Sus_scrofa_inverse))) #29
```


<i>

**Method notes:**

Take column belonging to each organism (will have all distances between that organism and all other organisms)


take inverse from all organisms and add them all up . single number is representation of histogram

How weighted are organisms towards low values. expect for mammals to have a common set of AMPs that fall outside of mammals (only distantly related to mammals) which will have low difference between them and to mammals. Values within mammals will change more (higher values in histogram)

set values to 0 for really distant taxonomic distance?


Sum up the inverse (1/distance) for distances so that a large value of distance translates to a small value (or weighting) of the AMP.
Inverse because an AMP that is taxonomically far away (e.g bacterial AMP to a human AMP) will be difficult to BLAST (not likely to have high sequence similarity) so we give that a low weighting. 

histogram for both normal and inverse?

</i>
